\ProjectName{}: Data Imputation as a Query Optimization

### 2016-12-14 paper/algorithm.tex
paper: some cleanup in language/typos

The computation of $\text{Time}(Q)$ is standard and is comparable to the cost model used in
any modern RDBMS.\@ Our runtime cost model is the cost model used in SimpleDB, though modern systems
have more sophisticated approaches.

$\text{Loss}(Q)$ is computed as follows.
For each sub-query, using the histogram information, we keep track of the estimated number of
null values in each column, $\text{MissingCount}(c, Q)$.
\begin{align*}
  \text{Loss}(Q) = \begin{cases}
    \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \delta_C(Q') \\
    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \mu_C(Q') \\
    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $p \in (0, 1]$ which decreases as the number of
tuples available to train the imputation algorithm increases.  Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information loss \textit{decreases} with more complete values to work
with. The inverse square loss can be seen as a heuristic first approximation to the learning guarantees of
regression trees. Since we conceive that
alternate imputation strategies could be used, the analyst would provide a corresponding
loss with a reasonable parameterization.

@@

\subsection{Query planning}
The input to our query planner is a tuple $(T, \Phi, \Psi, P, G, A)$: a set of tables $T$, a set of filter predicates $\phi_t, t \in T$, a set of join predicates $\psi_(t_1, t_2), t_1, t_2 \in T$, a set of attributes $P$, and an optional set of attributes $G$ and aggregation function $A \in \{\text{Max}, \text{Min}, \text{Sum}, \text{Avg}, \text{Count}\}$.

@@

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data} and others) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the analyst to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2010multiple}.

>>>>

Our runtime cost model is the cost model used in SimpleDB, though modern systems
have more sophisticated approaches.

For each sub-query, using the histogram information, we keep track of the estimated number of
null values in a column  ($\text{MissingCount}(c, Q)$). We use this estimate to compute
$\text{Loss}(Q)$ as follows.

\begin{align*}
  \text{Loss}(Q) = \begin{cases}
    \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \delta_C(Q') \\
    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \mu_C(Q') \\
    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $p \in (0, 1]$ which decreases as the number of
tuples available to train the imputation algorithm increases.  Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information loss \textit{decreases} with more complete values to work
with. The inverse square loss can be seen as a heuristic first approximation to the learning guarantees of
regression trees. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

@@

\subsection{Query planning}
The input to our query planner is a tuple $(T, \Phi, \Psi, P, G, A)$: a set of tables $T$, a set of filter predicates $\phi_t, t \in T$, a set of join predicates $\psi_{(t_1, t_2)}, t_1, t_2 \in T$, a set of attributes $P$, and an optional set of attributes $G$ and aggregation function $A \in \{\text{Max}, \text{Min}, \text{Sum}, \text{Avg}, \text{Count}\}$.

@@

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data}) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the analyst to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2010multiple}.

### 2016-12-14 paper/algorithms/histogram-transform.tex
paper: some cleanup in language/typos

%\begin{algorithm}
  \begin{algorithmic}
    \Require{$H$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
    \Ensure{Returns an updated $H$, such that the distribution of data matches the original $H$}

    \Function{UpdateHistogram}{$H, op$}
    	\If{$op \not\in \{\delta, \mu, \sigma, \bowtie \}$}
		\Return{H}
	\EndIf
	\State $H' \gets \Call{Copy}{H}$
    	\If{op = $\delta_C$}
		\For{$c \in C$}
			\State $H'[c][null] \gets 0$
		\EndFor
	\ElsIf{op = $\mu_C$}
		\For{$c \in C$}
			\State \Call{Add}{$H'[c], H'[c][null]$} 	\Comment{Adds to histogram based on distirbution}
		\EndFor
	\ElsIf{op = $\sigma$}
		\State \Call{ScaleBy}{$H',  \Call{Selectivity}{\sigma}$} 	\Comment{Scales all buckets by a constant factor}
	\ElsIf{op = $\bowtie_\phi$}
		\State \Call{ScaleTo}{$H', \Call{Cardinality}{\bowtie_{\phi}}$} 	\Comment{Scales all buckets to sum to constant}
	\EndIf
	\Return{H'}
    \EndFunction
  \end{algorithmic}
  \caption{An algorithm for in-plan histogram updates}
  \label{algo:histogram-transformation}
%\end{algorithm}

>>>>

%\begin{algorithm}
  \begin{algorithmic}
    \Require{$H$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
    \Ensure{Returns an updated histogram, such that the distribution of data matches the original $H$}

    \Function{UpdateHistogram}{$H, op$}
    	\If{$op \not\in \{\delta, \mu, \sigma, \bowtie \}$}
		\Return{H}
	\EndIf
	\State $H' \gets \Call{Copy}{H}$
    	\If{op = $\delta_C$}
		\For{$c \in C$}
			\State $H'[c][null] \gets 0$
		\EndFor
	\ElsIf{op = $\mu_C$}
		\For{$c \in C$}
			\State \Call{Add}{$H'[c], H'[c][null]$} 	\Comment{Adds to histogram based on distribution}
		\EndFor
	\ElsIf{op = $\sigma$}
		\State \Call{ScaleBy}{$H',  \Call{Selectivity}{\sigma}$} 	\Comment{Scales all buckets by constant factor}
	\ElsIf{op = $\bowtie_\phi$}
		\State \Call{ScaleTo}{$H', \Call{Cardinality}{\bowtie_{\phi}}$} 	\Comment{Scales all buckets to sum to constant}
	\EndIf
	\Return{$H'$}
    \EndFunction
  \end{algorithmic}
  \caption{An algorithm for in-plan histogram updates}
  \label{algo:histogram-transformation}
%\end{algorithm}

### 2016-12-14 paper/experiments.tex
paper: some cleanup in language/typos

In order to evaluate additional queries, we constructed a synthetic table, consisting of 10 columns and 10,000 rows, with
all values drawn from a uniform distribution between 0 and 100. To create a dirtied variant of the synthetic data, we randomly deleted 30\% of the fields
across the table.

@@

\item For average queries (\ref{q3}, \ref{q5}, \ref{q6}), imputation does not reduce the error, mainly because running these queries on dirty data incurs only a small penalty.
This is because, in the case of average, removing data uniformly at random does not change the mean.
If we were to remove data in a biased way, imputation would perform better because there would be more error for it to correct.

>>>>

In order to evaluate additional queries, we constructed a synthetic table, consisting of 10,000 rows and 10 columns, with
all values drawn from a uniform distribution between 0 and 100. To create a dirtied variant of the synthetic data, we randomly deleted 30\% of the fields
across the table.

@@

\item For average queries (\ref{q3}, \ref{q5}, \ref{q6}), imputation does not reduce the error, mainly because running these queries on dirty data incurs only a small penalty.
This is because, in the case of average, removing data uniformly at random does not change the mean.
If we were to remove data in a biased way, imputation would likely perform better because there would be more error for it to correct.

### 2016-12-14 paper/introduction.tex
paper: some cleanup in language/typos

Handling incorrect or dirty data is a complex and challenging problem for data scientists.
One way in which a dataset can be dirty is for parts of it to be missing altogether.
Missing data is one of the simpler variants of dirty data, but if handled naively, it can still cause analyses to be incorrect.
To handle this problem, users may manually clean their dataset by performing some regression analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
Although manual imputation solves the problem of missing data, in the age of big data, it may be very expensive to run an imputation algorithm on an entire dataset.
Additionally, it may not even be necessary to completely clean the data to make it usable.
Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
Others may want to run queries on a subset of the data, so do not need to impute every field of every record.

>>>>

Handling incorrect or dirty data is a complex and challenging problem for data scientists.
One way in which a dataset can be dirty is for parts of it to be missing altogether.
Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
To handle this problem, users may manually clean their dataset by performing some regression analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
Although manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
Additionally, it may not even be necessary to completely clean the data to make it usable.
Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
Others may want to run queries on a subset of the data, so do not need to impute every field in every record.

### 2016-12-14 paper/main.tex
start filling out algorithm section from onenote
paper: added join algo sketched, added chained equations algo sketched, added histogram transformation algo
paper: macro for project name
paper: switch to biblatex + biber
paper: add code to generate tables
paper: add to experiments section
paper: some cleanup, still missing motivating example and final experiments
paper: move related work
paper: keep working on experiment section
paper: some cleanup in language/typos

\documentclass{article}

\title{Data Imputation as a Query Optimization}

\author{
  Jose Cambronero \\
  \and
  Micah Cohen
  \and
  John K. Feser}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

@@

\begin{abstract}
In order to study the placement of an imputation step, we create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process in SimpleDB.  We introduce measures of uncertainty and completeness of query results as well as the cost of imputation, which outline the main trade-offs in the imputation placement. We add these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning. We show the trade-offs between efficiency and accuracy for simple data imputation models. \end{abstract}

\input{introduction}
\input{examples}
\input{algorithm}
\input{experiments}
\input{relatedwork}
\input{conclusion}

\bibliography{references}

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[style=ieee]{biblatex}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{hyperref} % load after other packages
\addbibresource{references.bib}

\makeatletter
\newcommand*{\centerfloat}{%
  \parindent \z@
  \leftskip \z@ \@plus 1fil \@minus \textwidth
  \rightskip\leftskip
  \parfillskip \z@skip}
\makeatother

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

\MakeRobust{\Call}

\renewcommand{\emptyset}{\varnothing}

\newcommand{\ProjectName}{ImputeDB}

\title{\ProjectName{}: Data Imputation as a Query Optimization}

\author{
  Jose Cambronero \\
  \and
  John K. Feser
  \and
  Micah Smith}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\newcommand{\todo}[1]{{\color{red}\textbf{TODO:} #1}}

@@

\begin{abstract}
In order to study the placement of an imputation step, we create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process in SimpleDB.  We introduce measures for information loss and runtime for imputation operations, which outline the main trade-offs in the imputation placement. We add these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning. We show the trade-offs between efficiency and accuracy for simple data imputation models. \end{abstract}

\input{introduction}
\input{relatedwork}
\input{examples}
\input{algorithm}
\input{experiments}
\input{conclusion}

\printbibliography

### 2016-12-14 paper/relatedwork.tex
paper: some cleanup in language/typos

\todo{Micah: Should this section follow the introduction? A lot of this stuff
becomes repetitive as we wouldn't be able to develop the body of the paper
without having already made these citations.}

@@

Database system developers and others have worked on techniques to automatically
detect dirty values, whether missing or otherwise, and rectify the errors is
possible. A survey of methods and systems is provided in
\cite{hellerstein2008quantitative}.

BayesDB \cite{mansinghka2015bayesdb} provides users with a simple interface to 
leverage statistical inference techniques on a database. Non-experts
can use a simple declarative language, akin to SQL, to specify models
which allow missing value imputation, amongst other broader functionality.
Experts can further customize strategies and express domain knowledge to
improve performance and accuracy.

>>>>

@@

Database system developers and others have worked on techniques to automatically
detect dirty values, whether missing or otherwise, and rectify the errors if
possible. A survey of methods and systems is provided in
\cite{hellerstein2008quantitative}.

BayesDB \cite{mansinghka2015bayesdb} provides users with a simple interface to 
leverage statistical inference techniques in a database. Non-experts
can use a simple declarative language, akin to SQL, to specify models
which allow missing value imputation, amongst other broader functionality.
Experts can further customize strategies and express domain knowledge to
improve performance and accuracy.
